---
title: '<font size = 7 color = "blue">HTML Presentations </font>'
subtitle: '<font size = 6 color = "blue">Data: x03 </font>'
author: '<font size = 6 color = "blue">Boning Liu </font>'
institute: '<!--img src = "https://www.wcupa.edu/communications/images/mainLogo1.png" width="130" height="100"--><br><font size = 6 color = "purple">West Chester University of Pennsylvania</font><br> '
date: '<font color = "blue" size =6> 09/25/2022 </font> <br> <font color = "blue" size = 6> STA 490: Capstone Statistics </font> <br> <br> <font color = "white" size = 5> '
output:
  xaringan::moon_reader:
    #yolo: true  #randomly insert new slides with background image of Karl’s mustache photo into existing slides
    lib_dir: libs
    self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
      #titleSlideClass: [center, middle, inverse]  # default set up of the title page
      #titleSlideClass: [top, left, inverse]
      #autoplay: 30000     # This sets autoplay, 30000 = 10 seconds
      #countdown: 30000    # The timer will count down to 00:00 then continue with red time
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
##
if(!require('xaringan')) {
  remotes::install_github("yihui/xaringan")
  library('xaringan')
}
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("leaflet")) {
   install.packages("leaflet")
   library(leaflet)
}
if (!require("ggplot2")) {
    install.packages("ggplot2")
    library(ggplot2)
}
if (!require("dplyr")) {
    install.packages("dplyr")
    library(dplyr)
}
if (!require("broom")) {
    install.packages("broom")
    library(broom)
}
if (!require("ggpubr")) {
    install.packages("ggpubr")
    library(ggpubr)
}
# remotes::install_github("yihui/xaringan")
```
class: middle

## <center>Data </center>
```{r}
x03 = read.table("https://people.sc.fsu.edu/~jburkardt/datasets/regression/x03.txt", skip = 36)
names(x03) = c("Index", "One", "Age", "Systolic Blood Pressure")
names(x03) = c("I", "A1", "A2", "B")
```

---
class: middle
## Data Table
```{r}
kable(x03)
```

---

## Histogram For Normality Check
```{r}
hist(x03$B)
```
---

## Building the Model
```{r}
Model <- lm(B ~ A2, data = x03)
summary(Model)
```
---
## Interpretation
From our model, we have a linear regression B = 0.97A2 + 98.71. The model has a multiple r-squared value of 0.432. This means that our model explains 43.2% of the variation within systolic blood pressure. This means that age helps to explain some of the variation, but not as much as we would like.
